---
title: "Data Processing Notebook"
output: rmarkdown
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(whowmr)
```

# WHO Annexes Data Processing Notes

This document serves as a workspace for ongoing notes compiled during
reviews of the WHO annexes. The purpose is twofold. Firstly, it is
difficult to say in advance how comparable the various datasets
contained within the annexes are across different years. Secondly, it is
anticipated that many users will require joining datasets from various
years to conduct analyses. Additionally, these notes aim to document
common data processing steps. For instance, a frequently encountered
column denoting "WHO Region/Country/Area" can be separated into distinct
columns for each element which is useful.

## Processing the `WHO Region/Country/Area` column

The `WHO Region/Country/Area` column is a common column in WHO datasets.
Usually the first row will give the name of the region, and the
following rows will give the names of countries in that region. The row
containing the region name will have no data in it. One exception to
this structure is observed with Tanzania, where the column is further
divided into `Mainland` and `Zanzibar`. In cases where Tanzania presents
data in all three cases, straightforward calculations of totals may lead
to duplicate counting of Tanzania's data due to this inclusion of
sub-regions.

A function is included in this package which attempts to split out the
`WHO Region/Country/Area` column in a dataset. Anecdotally it seems the
best approach is to label columns which are all upper-case characters as
"WHO Region" and all other columns as "Country/Area".

## Annex A: Policy logical columns

In 2017, Annex A gives policy columns with `Y` if true, `N` if false,
and either `-` or `NA`. There were no exceptions. The 2019:2023 reports
use multiple values and should not be boolean but rather a factor.

## Annex C: Funding

In reviewing the data, a misspelling was identified in Column G
"Government (NMP)" for the year 2017. Additionally, inconsistencies were
found in the terminology used between 2017 and 2018, where "NMCP" was
used instead of "NMP". Examination of data from 2019 onwards indicated
consistent use of "NMP", and considering that not all countries
participate in control programs, "NMP" was deemed the more accurate
term. To ensure consistency, the column names in 2017 and 2018 were
retrospectively changed to reflect this.

Furthermore, analysis of the 2019 data revealed the presence of
footnotes in the column following "Government (NMP)" which is different
to 2017/2018 where this column has only one possible value. To account
for the various potential interpretations of these footnotes, the column
name was subsequently changed to "Government Footnotes" (altering
backward compatibility of previous package versions). To reflect this
significant change, the major version number was incremented when
including 2019 data.

## Annex E - malaria indicator surveys from StatCompiler

This is the only Annex sheet without a "WHO Region/Country/area" column
owing to the fact that it grabs country data specifically.

The 2020 report contains a very useful sheet `annex_3e_with_links` which
has some resize-hidden columns/rows and formulas which show how exactly
how the calculations have been done. It is not difficult to use
StatCompiler to reproduce the analysis for all indicator surveys and all
countries with a single API call. A future version of this package could
include a function to get all surveys making this set of sheets
unnecessary. The above sheet would be very useful when writing such a
function.

### API Call

The indicators used seem mostly consistent between 2018:2023 with the
2017 containing a subset. Another exception is between 2018 and 2019
some indicators changed (shown in red in the 2019 annex Ea). There's
some further processing required (see Annex 3Ea in 2020 data) but the
same indicators seem to be used between reports. Manually putting in the
indicators used to generate the 2020 report in StatCompiler and
monitoring the API request gives the following:

```         
https://api.dhsprogram.com/rest/dhs/data/all?countryIds=AO,BU,ZW,ZM&indicatorIds=ML_NETP_H_ITN,ML_NETP_H_IT2,ML_IRSM_H_IRS,ML_IRSM_H_IIR,ML_IRSM_H_I2I,ML_ITNA_P_ACC,ML_NETU_P_ITN,ML_ITNU_N_ITN,ML_NETW_W_ITN,ML_IPTP_W_3SP,ML_NETC_C_ITN,CN_ANMC_C_MOD,CN_ANMC_C_SEV,ML_PMAL_C_RDT,ML_FEVT_C_ADV,ML_FEVT_C_BLD,ML_FEVT_C_AML,ML_AMLD_C_ACT&returnFields=DataId,Indicator,SurveyId,IsPreferred,Value,SDRID,Precision,RegionId,SurveyYearLabel,SurveyType,SurveyYear,IndicatorOrder,DHS_CountryCode,CILow,CountryName,CharacteristicId,CharacteristicCategory,IndicatorId,CharacteristicOrder,CharacteristicLabel,ByVariableLabel,DenominatorUnweighted,DenominatorWeighted,CIHigh,IsTotal,ByVariableId,IndicatorType,IndicatorTotalId&perpage=4000&apiKey=DHSPRG-949817&lang=en&page=2
```

These `indicatorIds` will likely be useful when interacting with the
API. The countries chosen seem to be iso2 codes. The
[rdhs](https://docs.ropensci.org/rdhs/) package would probably be a good
way to interact with the API. There are some package development ideas
worth exploring in there also.

### Rabbit hole: % of children \<5 years with fever in the past 2 weeks who took antimalarial drugs

Also note the calculation for "Children with fever who took antimalarial
drugs" since there are two variables "Three years preceding the survey"
and "Five years preceding the survey". In the 2020 sheet a `VLOOKUP` is
used which references a `Table2` which isn't a sheet but is included in
the Excel file. Exporting that table indicates that "Three years" comes
before the "Five years". That VLOOKUP asks for an Exact match (FALSE) so
the first value should be picked but somehow the "Five years" value is
being picked. It's probably best at this point to stop asking why and to
just make a note of it here in case someone wants to look into it
further for whatever reason.

### Other notes

The last column is under the major heading of "% of children \<5 years
with fever in the past 2 weeks" with the heading of "who took an ACT
among those who received any antimalarial". From sleuthing into the 2020
data it seems that this pulls the MIS indicator "Children who took any
ACT" whereas the previous indicators start with "Children with
fever...". Following up on this however, the documentation for the
denominator is "Children under age five with fever in the two weeks
preceding the survey that received any anti-malarial drugs" for this
indicator.

## Footnotes

There are many times where a cell will end in a number which links to a
footnote. While removing such numbers could have resulted in more
uniform and directly usable cells, the decision was made to generally
retain them in this package. This prioritizes user engagement with the
footnotes within the report, encouraging users to determine the
appropriate handling of the data.

For instance, the 2019 Annex 3C, covering funding information, includes
a footnote for "Venezuela" specifying that government funding values are
provided in local currency, not USD. Preserving this linked note
discourages users from assuming USD values during analysis, a potential
consequence of overlooking such crucial details.

## NA values in columns

It's most common to see "-" as meaning NA in a column. The footnotes
usually indicate how to interpret this.
